---
title: "GCRMN_Data_Tutorial"
author: "Noelle Helder"
date: "5/27/2024"
output: html_document
---
This tutorial will build on the skills we learned in the Data Vis tutorial. We will take it a step further, combining the steps into a function to automate this proces to work on multiple fish surveys at once! If you don't want to try and write your own functions and have cleaned your data in Excel, you can jump down to the 'Visualization' section to look at the `ggplot2` code. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# Define our Fish Data Cleaning Functions

The real power of programming for data analysis comes in when you have to repeat tasks multiple times. Doing this manually in Excel creates lots of opportunities for error. Copying and pasting data is a great way to create problems for yourself later. Programming out your data cleaning steps means that each step is well-documented so you always know what you did *without ever changing the original data*. This is key for **reproducible research**, because science is only valid to the extent that someone else could repeat it.

Last tutorial, we used a bunch of `functions` from packages in the `tidyverse` to clean our data. For example, we used the function `mutate` to change column names or `range` to identify the spread of a variable. 

```{r, echo=FALSE}
# Function to clean and reshape metadata
clean_and_reshape_metadata <- function(metadata, file_id) {
 
  m <- metadata[,1:2] %>%
    pivot_wider(names_from = ...1, values_from = ...2) %>% 
    mutate(ID = file_id)
  return(m)
}

# Function to clean and process data
clean_and_process_data <- function(data, file_id) {
  # Assign column names to the data
  col_names <- c("Species_Class", paste("Transect_", 1:5, sep = ""))
  colnames(data) <- col_names
  
  # Separate values in the first column into "Species" and "Class"
  data2 <- data %>%
    separate(Species_Class, into = c("Species", "Class"), sep = " ", remove = FALSE)
  
  # Convert columns to appropriate data types
  data2 <- data2 %>%
    mutate(across(starts_with("Transect_"), as.numeric)) %>%
    mutate(Species = as.factor(Species),
           Class = as.factor(Class))
  
  # Step 5: Reshape data to long format for visualization
  data_long <- data2 %>%
    pivot_longer(cols = starts_with("Transect_"), names_to = "Transect", values_to = "Count") %>%
    mutate(ID = file_id)
  
  return(data_long)
}

# Function to process each file
process_file <- function(file, file_id) {
  # Read the data
  df <- read_excel(file, col_names = FALSE)
  
  # Separate metadata and data
  metadata <- df[1:4, ]
  data <- df[-(1:5), ]
  
  # Clean and reshape metadata
  cleaned_metadata <- clean_and_reshape_metadata(metadata, file_id)
  
  # Clean and process data
  processed_data <- clean_and_process_data(data, file_id)
  
  return(list(metadata = cleaned_metadata, data = processed_data))
}

```



```{r}
# Set your working directory
setwd("C:/Users/nhelder/Documents/projects/Union_2024/GCRMN_data_vis")

file_paths <- list.files(path = "Fish_Data/", pattern = "*.xlsx", full.names = TRUE)


# Initialize empty lists to store the results
metadata_list <- list()
data_list <- list()

# Loop through each file, read the data, and apply the function
for (i in seq_along(file_paths)) {
  file <- file_paths[i]
  results <- process_file(file, file_id = i)
  
  metadata_list <- append(metadata_list, list(results$metadata))
  data_list <- append(data_list, list(results$data))
}

# Combine all the results into two separate dataframes
final_metadata_df <- bind_rows(metadata_list)
final_data_df <- bind_rows(data_list)

# View the final combined dataframes
print("Final Metadata Dataframe:")
print(final_metadata_df)

```

```{r}

print("Final Data Dataframe:")
print(final_data_df)
```
```{r}
df <- final_data_df %>% left_join(final_metadata_df)
df
write.csv(df, "fake_fish_surveys.csv", row.names = FALSE)
```

```{r}
ggplot(df) +
  aes(x=Species, 
      y=Count, 
      fill=Depth) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(~Site) +
  # add and edit lables, including titles and captions with labs(). 
  labs(x="Family", 
       y="Number Observed", 
       title="Title for my Very Important Data",
       subtitle = "Subtitle for my very important fish count data") +
  # add a generic theme
  theme_bw() + 
  # modify specific parts of the theme
  theme(legend.position="bottom", # move the legend below the plot
        # get rid of the legend title
        # legend.title=element_blank(),
        # get rid of the x-axis title
        axis.title.x=element_blank(),
        # get rid of the x-axis labels
        # axis.text.x=element_blank(),
        # get rid of the x-axis tick marks
        axis.ticks.x=element_blank())
```

```{r}
# use summarize to calculate the average for each transect
df$Count[is.na(df$Count)] <- 0
summary <- df %>% 
  group_by(Species, Site, Depth) %>% 
  summarize(avg=mean(Count),
            sd = sd(Count))
summary
```
I can specify colors for grouping variables for consistency. 
```{r}
group.colors <- c('15 m' = "#66E6FF", '30 m'= "#007A99")
```

```{r}
ggplot(summary) +
  aes(x=Species, 
      y=avg, 
      fill=Depth) +
  geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin =avg, ymax=avg+sd), width=0.1,  position=position_dodge(.9)) +
 #  geom_errorbar(aes(ymin = avg, ymax = avg + sd), 
  #               width = 0.2) +  # Add error bars
  facet_grid(~Site) +
  # add and edit lables, including titles and captions with labs(). 
  labs(x="Family", 
       y="Abundance", 
       title="Title for my Very Important Data",
       subtitle = "Subtitle for my very important fish count data") +
  # add a generic theme
  theme_classic() + 
  scale_fill_manual(values=group.colors) +
  # modify specific parts of the theme
  theme(legend.position="bottom", # move the legend below the plot
        # get rid of the x-axis title
        axis.title.x=element_blank(),
        # Change x-axis label size so that it all fits
        axis.text.x =element_text(size = 8),   
        # get rid of the x-axis tick marks
        axis.ticks.x=element_blank())

```